---
title: Clustering and classification
author: "Oyedayo Oyelowo"
date: "26 November 2017"
output: html_document
---

```{r warning=FALSE, message=FALSE}
# MASS package
library(MASS)
library(ggplot2)
library(GGally)
library(tidyr)
#install.packages("plotly")
library(tidyverse)
library(corrplot)
library(dplyr)
library(plotly)
data("Boston")

#structure of the data
str(Boston)

#dimension of the data
dim(Boston)

#Describe the data

# Boston {MASS}	R Documentation
# Housing Values in Suburbs of Boston
# 
# Description
# 
# The Boston data frame has 506 rows and 14 columns.
# 
# Usage
# 
# Boston
# Format
# 
# This data frame contains the following columns:
# 
# crim
# per capita crime rate by town.
# 
# zn
# proportion of residential land zoned for lots over 25,000 sq.ft.
# 
# indus
# proportion of non-retail business acres per town.
# 
# chas
# Charles River dummy variable (= 1 if tract bounds river; 0 otherwise).
# 
# nox
# nitrogen oxides concentration (parts per 10 million).
# 
# rm
# average number of rooms per dwelling.
# 
# age
# proportion of owner-occupied units built prior to 1940.
# 
# dis
# weighted mean of distances to five Boston employment centres.
# 
# rad
# index of accessibility to radial highways.
# 
# tax
# full-value property-tax rate per \$10,000.
# 
# ptratio
# pupil-teacher ratio by town.
# 
# black
# 1000(Bk - 0.63)^2 where Bk is the proportion of blacks by town.
# 
# lstat
# lower status of the population (percent).
# 
# medv
# median value of owner-occupied homes in \$1000s.
# 
# Source
# 
# Harrison, D. and Rubinfeld, D.L. (1978) Hedonic prices and the demand for clean air. J. Environ. Economics and Management 5, 81-102.
# 
# Belsley D.A., Kuh, E. and Welsch, R.E. (1980) Regression Diagnostics. Identifying Influential Data and Sources of Collinearity. New York: Wiley.
```


Show a graphical overview of the data and show summaries of the variables in the data. Describe and interpret the outputs, commenting on the distributions of the variables and the relationships between them. (0-2 points)
```{r warning=FALSE, message=FALSE}
#graphical overview of the data
#ggpairs(Boston)
#pairs(Boston)
#plot(Boston)

#summary of the data
summary(Boston)

# calculate the correlation matrix and round it
cor_matrix<-cor(Boston)


# visualize the correlation matrix
corrplot(cor_matrix, method="circle")





```


Standardize the dataset and print out summaries of the scaled data. How did the variables change? Create a categorical variable of the crime rate in the Boston dataset (from the scaled crime rate). Use the quantiles as the break points in the categorical variable. Drop the old crime rate variable from the dataset. Divide the dataset to train and test sets, so that 80% of the data belongs to the train set. (0-2 points)
```{r warning=FALSE, message=FALSE}
# MASS and Boston dataset are available

# center and standardize variables
boston_scaled <- scale(Boston)

# summaries of the scaled variables
summary(boston_scaled)

# class of the boston_scaled object
class(boston_scaled)

# change the object to data frame
boston_scaled<-as.data.frame(boston_scaled)


# MASS, Boston and boston_scaled are available

# summary of the scaled crime rate
summary(boston_scaled$crim)

# create a quantile vector of crim and print it
bins <- quantile(boston_scaled$crim)
bins

# create a categorical variable 'crime'
crime <- cut(boston_scaled$crim, breaks = bins, include.lowest = TRUE, label=c("low", "med_low", "med_high", "high"))

# look at the table of the new factor crime
table(crime)

# remove original crim from the dataset
boston_scaled <- dplyr::select(boston_scaled, -crim)

# add the new categorical value to scaled data
boston_scaled <- data.frame(boston_scaled, crime)


# boston_scaled is available

# number of rows in the Boston dataset 
n <- nrow(boston_scaled)

# choose randomly 80% of the rows
ind <- sample(n,  size = n * 0.8)

# create train set
train <- boston_scaled[ind,]

# create test set 
test <- boston_scaled[-ind,]

# save the correct classes from test data
correct_classes <- test[,"crime"]

# remove the crime variable from test data
test <- dplyr::select(test, -crime)


# MASS and train are available

# linear discriminant analysis
lda.fit <- lda(crime~., data = train)

# print the lda.fit object
lda.fit

# the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "red", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

# target classes as numeric
train$crime <- as.numeric(train$crime)

# plot the lda results
plot(lda.fit, dimen = 2, col = train$crime, pch= train$crime)
lda.arrows(lda.fit, myscale = 2)
```
```{r warning=FALSE, message=FALSE}
# predict classes with test data
lda.pred <- predict(lda.fit, newdata = test)

# cross tabulate the results
table(correct = correct_classes , predicted = lda.pred$class )
```

Reload the Boston dataset and standardize the dataset (we did not do this in the Datacamp exercises, but you should scale the variables to get comparable distances). Calculate the distances between the observations. Run k-means algorithm on the dataset. Investigate what is the optimal number of clusters and run the algorithm again. Visualize the clusters (for example with the pairs() or ggpairs() functions, where the clusters are separated with colors) and interpret the results. (0-4 points)


```{r warning=FALSE, message=FALSE}
# load MASS and Boston
library(MASS)
data('Boston')

boston_standard <- scale(Boston)
# euclidean distance matrix
dist_eu <-dist(boston_standard)

# look at the summary of the distances
summary(dist_eu)

# manhattan distance matrix
dist_man <- dist(boston_standard, method="manhattan")

# look at the summary of the distances
summary(dist_man)


# Boston dataset is available

# k-means clustering
km <-kmeans(Boston, centers = 3)

# plot the Boston dataset with clusters
pairs(Boston[6:8], col = km$cluster)



set.seed(123)

# determine the number of clusters
k_max <- 10

# calculate the total within sum of squares
twcss <- sapply(1:k_max, function(k){kmeans(Boston, k)$tot.withinss})

# visualize the results
qplot(x = 1:k_max, y = twcss, geom = 'line')

# k-means clustering
km <-kmeans(Boston, centers = 2)

# plot the Boston dataset with clusters
pairs(Boston[5:7], col = km$cluster)


```


Bonus: Perform k-means on the original Boston data with some reasonable number of clusters (> 2). Remember to standardize the dataset. Then perform LDA using the clusters as target classes. Include all the variables in the Boston data in the LDA model. Visualize the results with a biplot (include arrows representing the relationships of the original variables to the LDA solution). Interpret the results. Which variables are the most influencial linear separators for the clusters? (0-2 points to compensate any loss of points from the above exercises)


```{r warning=FALSE, message=FALSE}
boston_standard2<-scale(Boston)
set.seed(123)

# determine the number of clusters
k_max <- 20

# calculate the total within sum of squares
twcss <- sapply(1:k_max, function(k){kmeans(boston_standard2, k)$tot.withinss})

# visualize the results
qplot(x = 1:k_max, y = twcss, geom = 'line')

# k-means clustering
km <-kmeans(boston_standard2, centers = 9)

#convert km to dataframe
boston_standard2<-as.data.frame(boston_standard2)

lda.fit_clus<- lda(km$cluster~., data=boston_standard2)

# plot the Boston dataset with clusters
pairs(boston_standard2[,3:7], col = km$cluster)

# the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "red", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

# target classes as numeric
classes <- as.numeric(train$crime)

# plot the lda results
plot(lda.fit_clus, dimen = 2, col = classes, pch= classes)
lda.arrows(lda.fit, myscale = 2)
```


Super-Bonus: Run the code below for the (scaled) train data that you used to fit the LDA. The code creates a matrix product, which is a projection of the data points.

```{r warning=FALSE, message=FALSE}
model_predictors <- dplyr::select(train, -crime)
# check the dimensions
dim(model_predictors)
dim(lda.fit$scaling)
# matrix multiplication
matrix_product <- as.matrix(model_predictors) %*% lda.fit$scaling
matrix_product <- as.data.frame(matrix_product)

#Next, install and access the plotly package. Create a 3D plot (Cool!) of the #columns of the matrix product by typing the code below.

plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type= 'scatter3d', mode='markers')


# Adjust the code: add argument color as a argument in the plot_ly() function. Set the color to be the crime classes of the train set. Draw another 3D plot where the color is defined by the clusters of the k-means. How do the plots differ? Are there any similarities? (0-3 points to compensate any loss of points from the above exercises)

plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type= 'scatter3d', mode='markers', col=classes)


plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type= 'scatter3d', mode='markers', col=km$cluster)
```
