
#Logistic regression

The data includes students' secondary education accomplishment of two portuguese schools. The data attributes include student grades, demographic, social and school related features')  and it was collected by using school reports and questionnaires. Two datasets are provided regarding the performance in two distinct subjects: Mathematics (mat) and Portuguese language (por). In [Cortez and Silva, 2008], the two datasets were modeled under binary/five-level classification and regression tasks. Important note: the target attribute G3 has a strong correlation with attributes G2 and G1. This occurs because G3 is the final year grade (issued at the 3rd period), while G1 and G2 correspond to the 1st and 2nd period grades. It is more difficult to predict G3 without G2 and G1, but such prediction is much more useful (see paper source for more details).
(This is the data source)[https://archive.ics.uci.edu/ml/datasets/Student+Performance]


###Data Set Information:
Data was gotten from this (source) [http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/alc.txt] 
  

```{r, message=FALSE, warning=FALSE}
# access the tidyverse libraries tidyr, dplyr, ggplot2
#install.packages("tidyr")
library(tidyr); library(dplyr); library(ggplot2); library(GGally);

#read the data
alc<- read.table("http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/alc.txt", sep = ",", header = T)
```

###DATA OVERVIEW
```{r, message=FALSE, warning=FALSE}
colnames(alc)   #names of columns
summary(alc)    #summary
dim(alc)    #dimension
str(alc)   #structure
```


```{r, message=FALSE, warning=FALSE}
attach(alc)   #attach all the variables in the dataframe
# define a new logical column 'high_use'
alc$high_use<- alc$alc_use > 2
#alc <- mutate(alc, high_use = alc_use > 2) #alternative
```

###Hypotheses.
-  _age is not related to alcohol consumption_
-  _sex is not related to alcohol consumption_
-  _absence from school is not related to alcóhol use_
-  _freetime after school is not related to alcohol consumption_


##subsetting and exploring my chosen variables
```{r, message=FALSE, warning=FALSE}
hyp<- alc[,c("age", "sex",  "absences","freetime","alc_use")]
#exploring my chosen variables
# draw a bar plot of each variable
hyp<-gather(hyp) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free")
hyp + geom_bar()
```
The above shows the distribution of the chosen predictors and also scale of alcohol consumption. The ages of the students are mainly within 15-19 with few students aged 20 and 22. Generally, there are relatively few chronic alcohol consumers amongst the students.
The students seem to have quite enough free time. The respondents include 198 female students and 184 male students.


##Overview of the chosen predictors and alcohol consumption variable.
```{r, message=FALSE, warning=FALSE}

hyp<- alc[,c("age", "sex",  "absences","freetime","alc_use")]
#show the overview of the predictors and response variable(non-binomial)
plot_hyp <- ggpairs(hyp, mapping = aes(col=sex, alpha=0.3), lower = list(combo = wrap("facethist", bins = 20)))
plot_hyp   #draw the plot
```
Here, we can see that the female students seem to be generally quite older and there are both male and female students that are much older than the general sample of the students.  The absences are not so high but some chronic absentees amongst the students(both male and female). Most of the students are quite free but a few of the students are very busy.
Largely, all the predictors show no correlation, hence, the issue of multicolinearity is not a problem here. However, these predictors still show very weak correlation  with alcohol consumption. 
Nevertheless, it should be recalled that correlation is not causation. Hence, I will dig further to understand the effects of these predictors on alcohol consumption amongst students and see if they are significant.


##Crosstabs
**The below are crosstabs that compare the predictors with alcohol consumption**
```{r, message=FALSE, warning=FALSE}
#using the binomial response(high_use)
#it is also possible to use simple crosstabs e.g
#xtabs(high_use~age, data = alc)
#but below is a more comprehensive crosstab.
#Crosstabs
summarise(group_by(alc, age,high_use), count=n(), mean_grade=mean(G3))
#an alternative and preferrable way of doing the above
#alc %>% group_by(age, high_use) %>% summarise(count=n(), mean_grade =  mean(G3))

summarise(group_by(alc, sex,high_use), count=n(), mean_grade=mean(G3))
summarise(group_by(alc, absences,high_use), count=n(), mean_grade=mean(G3))
summarise(group_by(alc, freetime,high_use), count=n(), mean_grade=mean(G3))
```


##Bar chats
###The age distribution by sex
```{r, message=FALSE, warning=FALSE}
# initialize a plot of 'age'
sex_age <- ggplot(data = alc, aes(x=age, col=sex))

# draw a bar plot of age by sex
sex_age + geom_bar() + facet_wrap("sex")
```


###high alcohol consumption by sex
```{r, message=FALSE, warning=FALSE}
# initialize a plot of 'high_use'
hu_sex <- ggplot(data = alc, aes(x=sex, col=sex))

# draw a bar plot of high_use by sex
hu_sex + geom_bar() + facet_wrap("high_use")
```
Here, we can see that there are more female who do are not high consumers of alcohol compared to men. Men consume more alchols than their female counterparts.


###High use vs freetime
```{r, message=FALSE, warning=FALSE}
# initialize a plot of 'high_use'
hu_ft <- ggplot(data = alc, aes(x=freetime, col=sex))

# draw a bar plot of 'high_use' by freetime
hu_ft + geom_bar() + facet_wrap("high_use")
```
Most students with more freetime seem not to consume alcohol as one would expect


###High use vs absences
```{r, message=FALSE, warning=FALSE}
# initialize a plot of 'high_use'
hu_ab <- ggplot(data = alc, aes(x=absences, col=sex))

# draw a bar plot of 'high_use' by freetime
hu_ab + geom_bar() + facet_wrap("high_use")
#few of the absentees are alcoholic but many are not.
```

###High use vs age
```{r, message=FALSE, warning=FALSE}
# initialize a plot of 'high_use'
hu_ag <- ggplot(data = alc, aes(x=age, col=sex))

# draw a bar plot of 'high_use' by freetime
hu_ag + geom_bar() + facet_wrap("high_use")
```
some of the young students below 20 years are high consumers of alcohol, but less so of those that are not high alchol consumers.


###High use vs romantic
```{r, message=FALSE, warning=FALSE}
# initialize a plot of 'high_use'
hu_rom <- ggplot(data = alc, aes(x=romantic, col=sex))

# draw a bar plot of 'high_use' by freetime
hu_rom + geom_bar() + facet_wrap("high_use")
```
For students in romantic relationships, they generally have lower number of them that are high consumers of alcohol. This is same with those in no romatic relationships.


##BOXPLOTS
**Below are boxplots to give clearer pictures, as explained earlier.
##Relationship of alcohol consumption with absences
```{r, message=FALSE, warning=FALSE}
# initialise a plot of high_use and absences
h_ab<- ggplot(alc, aes(x=high_use, y=absences,col=sex))

# define the plot as a boxplot and draw it
h_ab + geom_boxplot() + ggtitle("Student absences vs alcohol consumption")
```
There are a few chronic absentees amongst male and female students, but generally, most of the students have low absences.


###Relationship of alcohol consumption with age
```{r, message=FALSE, warning=FALSE}
# initialise a plot of high_use and age
h_ag<- ggplot(alc, aes(x=high_use, y=age,col=sex))

# define the plot as a boxplot and draw it
h_ag + geom_boxplot() + ggtitle("Students' age vs alcohol consumption")
```


###Relationship of alcohol consumption with freetime
```{r, message=FALSE, warning=FALSE}
# initialise a plot of high_use and freetime
h_fr<- ggplot(alc, aes(x=high_use, y=freetime,col=sex))

# define the plot as a boxplot and draw it
h_fr + geom_boxplot() + ggtitle("Student's freetime vs alcohol consumption")
```



###Relationship of alcohol consumption with sex
```{r, message=FALSE, warning=FALSE}
# initialise a plot of high_use and sex
alc_sex<- ggplot(alc, aes(y=alc_use, x=sex,col=sex))

# define the plot as a boxplot and draw it
alc_sex + geom_boxplot() + ggtitle("Student's alcohol consumption by sex")
```
There are a few female students with quite high alcohol consumption but the alcohol consumption levels do not vary as much as they do amongst the male students.




##Logistic regression

###fitting the glm model
```{r, message=FALSE, warning=FALSE}
high_use_mod1<- glm(high_use~ age + sex + absences + freetime , data= alc, family = "binomial")
summary(high_use_mod1)
# coef(high_use_mod1)
# coefficients(high_use_mod1)
```
The predictors "age" and "freetime "are not significant. Hence, these could be  considered as redundant variables and not have consierable effect on alcohol consumption compared with asences and sex which are significant.
hence, I could accept the null hypothesis that age and freetime are not 
related to alcohol consumption.
On the other hand, I reject the null hypothesis that male sex and absences are not related to alcohol consumption. They both have positive effects on alcohol consumption.

###Refitting the model without the redundant variables
```{r, message=FALSE, warning=FALSE}
high_use_mod2<- glm(high_use~  sex + absences, data= alc, family = "binomial")
summary(high_use_mod2)
#options("contrasts")
```

**check the overall effect of the variable by performing a likelihood ratio test**
```{r, message=FALSE, warning=FALSE}
anova(high_use_mod1, high_use_mod2, test="LRT")
```
There seems to be no significant difference between the two models.  
Hence, 'sex' and 'absences' are significant enough without the redundant variables.


**let's also see if "sex" is alos significant in the prediction**
```{r, message=FALSE, warning=FALSE}
high_use_mod3<- glm(high_use~  absences, data= alc, family = "binomial")
anova(high_use_mod2, high_use_mod3, test="LRT")
```
The test shows that sex is important and would remain in the model.

Hence, final model is: **_high_use~  sex + absences_**

###calculating the odds ratio and confident interval
```{r, message=FALSE, warning=FALSE}
# compute odds ratios (odds_ra)
odds_ra <- exp(coef(high_use_mod2))
#odds_ra <- coef(high_use_mod2) %>% exp     #alternaive

# compute confidence intervals (conf_int)
conf_int <- exp(confint(high_use_mod2)) 
#Conf_Int <- high_use_mod2 %>%  confint() %>% exp   #alternative

# print out the odds ratios with their confidence intervals
cbind(odds_ra, conf_int)
```
Here, we can see that Male sex and absences are positively associated with
success and are more likely to affect alcohol consumption. The confident interval of Male sex is quite high compared to "absences" which is narrower and much more likely to affect alcohol consumption.


###fit the model
```{r, message=FALSE, warning=FALSE}
# predict() the probability of high_use
probs<- predict(high_use_mod2, type = "response")
```

**add the predicted probabilities to 'alc'**
```{r, message=FALSE, warning=FALSE}
alc$prob_high_use <- probs
#alc <- mutate(alc, probability = probabilities)
```

**use the probabilities to make a prediction of high_use, setting 0.5 as threshold**
```{r, message=FALSE, warning=FALSE}
alc$predict_high_use<- (alc$prob_high_use)>0.5
#alc <- mutate(alc, prediction = prob_high_use>0.5)
```
**see the first ten and last ten original classes, predicted probabilities, and class predictions**
```{r, message=FALSE, warning=FALSE}
head(alc[,c("failures", "absences", "sex", "high_use", "prob_high_use", "predict_high_use")], n=10)
select(alc, failures, absences, sex, high_use, prob_high_use, predict_high_use) %>% tail(10)
```

### tabulate the target variable versus the predictions
```{r, message=FALSE, warning=FALSE}
table(high_use = alc$high_use, prediction = alc$predict_high_use)
```
The model rightly predicted 263 False high use and 23 True high_use of alcohol.
It wrongly predicted 89 True high_use and 7 False high_use

###Plotting the predicted and observed alcohol consumption
```{r, message=FALSE, warning=FALSE}
# initialize a plot of 'high_use' versus 'probability' in 'alc'
g <- ggplot(alc, aes(x = prob_high_use, y = high_use, col= predict_high_use))

# define the geom as points and draw the plot
g + geom_point()
```
#The wrong preditctions were quite much

### tabulate the target variable versus the predictions
```{r, message=FALSE, warning=FALSE}
conf_mat<-table(high_use = alc$high_use, prediction = alc$predict_high_use)
conf_mat<-prop.table(conf_mat)
addmargins(conf_mat)

#Alternatively, this can be done as shown below:
#addmargins(prop.table(table(high_use = alc$high_use, prediction = alc$predict_high_use)))
#table(high_use = alc$high_use, prediction = alc$predict_high_use) %>%  prop.table() %>% addmargins()
```


###mean error of the prediction
```{r, message=FALSE, warning=FALSE}
mean(abs(alc$high_use-alc$prob_high_use)>0.5)
```
My model has slightly lower error  of 0.25 compared with the one on datacamp with 0.26 mean error.


**The below is an alternative way by firstly defining the function to calculate the mean error.**
```{r, message=FALSE, warning=FALSE}
# define a loss function (mean prediction error)
loss_func <- function(class, prob) {
  n_wrong <- abs(class - prob) > 0.5
  mean(n_wrong)
}

```

```{r, message=FALSE, warning=FALSE}
# call loss_func to compute the average number of wrong predictions in the (training) data
loss_func(class = alc$high_use, prob = alc$prob_high_use)
```


### K-fold cross-validation
```{r, message=FALSE, warning=FALSE}
library(boot)
cv <- cv.glm(data = alc, cost = loss_func, glmfit = high_use_mod2, K = 10)

# average number of wrong predictions in the cross validation
cv$delta[1]
```
